---
title: "Data analysis excluding Chichinautzin gemorphological unit"
author: Sebastian Block
output:
  html_document:
    keep_md: true
    theme: flatly
---

This report shows an additional analysis presented in the paper Block and Meave (2017), published in the journal Plant Ecology & Diversity.  
The analysis was reproduced for the last time on May 12th, 2017. I used R version 3.3.0 (2016-05-03) -- "Supposedly Educational" and worked in a platform x86_64-apple-darwin13.4.0 (64-bit).

## Redundancy analysis

First, load the packages required for the analysis.

```{r}
library(ade4)
library(vegan)
library(MASS)
library(ellipse)
library(FactoMineR)
```

To load the data, all files must be in the working directory. Since the Chichinautzin geomorphological unit has extreme values of exposed rock compared to all other units, it is possible that this unit alone is driving the effects of rock exposure. Hence, I also did the analysis excluding data from this unit.


```{r}
structure <- read.csv("str.csv")
env <- read.csv("env.csv")
asp <- read.csv("aspect.csv")

# Remove Chichinautzin unit
ch.index <- env$unit != 'CH'
structure <- structure[ch.index,]
asp <- asp[ch.index,]
env <- env[ch.index,]

env$unit <- factor(env$unit)
par(mar = c(4.8, 4.8, 1, 1))
boxplot(rock~ unit, data = env, xlab = "Geomorphological unit", ylab = "% of exposed rock", cex.lab = 1.3)
```

I did some modifications to the data. First, I combined all the quantitative variables of the `env` data (i.e., all except the geomorphological unit) with the `asp` data into a single object called `env.quant`. Then, I normalized the data in `env.quant` to allow comparison between variables with very different units. Finally, I added the geomorphological unit column to the normalized quantitative environmental variables to create the data frame `env.ra`, which was used for the redundancy analysis. 

```{r}
env.quant <- cbind(env[,-5],asp)
env.z <- scale(env.quant)
unit <- factor(env[,5])
env.ra <- data.frame(env.z,unit)
```

### Structural data

I did a redundancy analysis on vegetation structure using the environmental data `env.ra` as explanatory variables. I assigned the results to the object `str.rda`.

```{r}
str.rda <- rda(structure~., env.ra, scale=T)
summary(str.rda)
```

I then got the R squared and the adjusted R squared for this analysis.

```{r}
(R2 <- RsquareAdj(str.rda)$r.squared)
(R2adj <- RsquareAdj(str.rda)$adj.r.squared)
```

To test the null hypothesis that no linear relationship exists between the response data and the explanatory variables, I did a permutation analysis with 1000 permutations. 

```{r}
anova.cca(str.rda, step=1000)
```

To make a triplot with scaling 2. 

```{r}
par(mar=c(4,4,2,2))
plot(str.rda, xlab="RDA1 (11.6 %)", ylab="RDA2 (3.9 %)", 
     display=c("cn", "lc", "sp"), type="n", xlim=c(-1.3, 1.3))
sites.sc <- scores(str.rda, choices=1:2, scaling=2, display="lc")
points(sites.sc, pch=1, cex=0.5)
va.sc <- scores(str.rda, choices=1:2, scaling=2, display="sp")
text(va.sc, row.names(va.sc), cex=0.9)
env.sc <- scores(str.rda, choices=1:2, scaling=2, display="bp")
arrows(0,0, env.sc[1:5,1], env.sc[1:5,2], lty=1, lwd=2.5, length=0.1)
env.names <- c("Elevation", "Slope", "Rock", "CDI", "Aspect")
text(env.sc[c(1,3,4),], env.names[c(1,3,4)], cex=0.9, font=2, pos=4)
text(env.sc[c(2,5),], env.names[c(2,5)], cex=0.9, font=2, pos=2)
unit.names <- c("LO", "OC", "SU", "TE", "UO")
unit.sc <- scores(str.rda, choices=1:2, scaling=2, display="cn")
points(unit.sc, pch=23)
text(unit.sc, unit.names, cex=0.7, font=2, pos=4)
#text(unit.sc[c(3),], unit.names[c(3)], cex=0.7, font=2, pos=3)
```

### Composition data

#### Canopy

First I loaded the species data. Again, the file `can.csv` must be in the working directory. To reduce the importance of large abundances, I applied the Hellinger transformation to the species data, as recommended by Borcard et al. (2011, Numerical Ecology in R)
. 
```{r}
can <- read.csv("can.csv", sep=",", header=T, row.names=1)
can <- can[ch.index,]
can.he <- decostand(can,"hellinger")
```

Then I did a redundancy analysis using all the environmental variables as explanatory variables. I tested the null hypothesis with a permutation analysis.

```{r}
can.rda <- rda(can.he~., env.ra)
summary.canrda <- summary(can.rda)
head(summary.canrda)

# Get the R^2 and an adjusted R^2
(spR2 <- RsquareAdj(can.rda)$r.squared)
(spR2adj <- RsquareAdj(can.rda)$adj.r.squared)

# Permutation analysis
anova.cca(can.rda, step=1000) 
```

To make a triplot of the RDA on canopy species data with scaling 2.

```{r}
# Make plotting space
par(mar=c(4,4,2,2))
plot(can.rda, scaling=2, display=c("cn", "lc", "sp"), type="n", 
     xlab="RDA1 (16.68 %)", ylab="RDA2 (7.96 %)", main = "Canopy")

# Plot points for sites
spsites.sc <- scores(can.rda, choices=1:2, scaling=2, display="lc")
points(spsites.sc, pch=1, cex=0.5)

# Plot points for species
sp.sc <- scores(can.rda, choices=1:2, scaling=2, display="sp")
points(sp.sc, pch=4, cex=0.5, col="gray50")

# Plot point for geomorphological unit centroids and their labels
unit.names <- c("LO", "OC", "SU", "TE", "UO")
canunit.sc <- scores(can.rda, choices=1:2, scaling=2, display="cn")
points(canunit.sc, pch=23, bg = "darkgray")
text(canunit.sc, unit.names, cex=0.7, font=2, pos=2, col = "darkgray")
#text(canunit.sc[c(3,6),], unit.names[c(3,6)], cex=0.7, font=2, pos=3, col = "darkgray")
#text(matrix(canunit.sc[1,], nrow=1, ncol=2), unit.names[c(1)], cex=0.7, font=2, pos=1, col = "darkgray")

# Plot arrows of quantitative explanatory variables and their labels
canenv.sc <- scores(can.rda, choices=1:2, scaling=2, display="bp")
arrows(0,0, canenv.sc[1:5,1], canenv.sc[1:5,2], lty=1, lwd=2.5, length=0.1, col="black")
env.names <- c("Elevation", "Slope", "Rock", "CDI", "Aspect")
text(matrix(canenv.sc[1,], nrow = 1, ncol = 2), env.names[1], cex=0.9, font=2, pos=2)
text(matrix(canenv.sc[3,], nrow = 1, ncol = 2), env.names[3], cex=0.9, font=2, pos=1)
text(canenv.sc[c(2,4),], env.names[c(2,4)], cex=0.9, font=2, pos=3)
text(canenv.sc[5,], env.names[5], cex=0.9, font=2, pos=4)
text(matrix(canenv.sc[5,], nrow = 1, ncol = 2), env.names[5], cex=0.9, font=2, pos=4)
```


#### Understory

First I loaded the species data. Again, the file `und.csv` must be in the working directory. To reduce the importance of large abundances, I applied the Hellinger transformation to the species data, as recommended by Borcard et al. (2011, Numerical Ecology in R)
. 
```{r}
und <- read.csv("und.csv", sep=",", header=T)
und <- und[ch.index,]
und.he <- decostand(und, "hellinger")
```

Then I did a redundancy analysis using all the environmental variables as explanatory variables. I tested the null hypothesis with a permutation analysis.

```{r}
und.rda <- rda(und.he ~ ., env.ra)
summary.undrda <- summary(und.rda)
head(summary.undrda)

# Get the R^2 and an adjusted R^2
(spR2 <- RsquareAdj(und.rda)$r.squared)
(spR2adj <- RsquareAdj(und.rda)$adj.r.squared)

# Permutation analysis
anova.cca(und.rda, step=1000) 
```

To make a triplot of the RDA on species data with scaling 2.

```{r}
# Make plotting space
par(mar=c(4,4,2,2))
plot(und.rda, scaling=2, display=c("cn", "lc", "sp"), type="n", 
     xlab="RDA1 (8.22 %)", ylab="RDA2 (7.17 %)", main = "Understory")

# Plot points for sites
spsites.sc <- scores(und.rda, choices=1:2, scaling=2, display="lc")
points(spsites.sc, pch=1, cex=0.5)

# Plot points for species
sp.sc <- scores(und.rda, choices=1:2, scaling=2, display="sp")
points(sp.sc, pch=4, cex=0.5, col="gray50")

# Plot point for geomorphological unit centroids and their labels
unit.names <- c("LO", "OC", "SU", "TE", "UO")
spunit.sc <- scores(und.rda, choices=1:2, scaling=2, display="cn")
points(spunit.sc, pch=23, bg = "darkgray")
text(spunit.sc, unit.names, cex=0.7, font=2, pos=2)
#text(spunit.sc[c(1,4,5,6),], unit.names[c(1,4,5,6)], cex=0.7, font=2, pos=4)

# Plot arrows of quantitative explanatory variables and their labels
spenv.sc <- scores(und.rda, choices=1:2, scaling=2, display="bp")
arrows(0,0, spenv.sc[1:5,1], spenv.sc[1:5,2], lty=1, lwd=2.5, length=0.1, col="black")
env.names <- c("Elevation", "Slope", "Rock", "CDI", "Aspect")
text(spenv.sc[c(1,3,4),], env.names[c(1,3,4)], cex=0.9, font=2, pos=1)
text(matrix(spenv.sc[2,], nrow = 1, ncol = 2), env.names[2], cex=0.9, font=2, pos=4)
text(matrix(spenv.sc[5,], nrow = 1, ncol = 2), env.names[5], cex=0.9, font=2, pos=2)
```


### Plot for the paper

This is the code to make the figure for the paper.

```{r, fig.width=6.7, fig.height=10}
par(mfrow=c(2,1))

# Figure 4a - Canopy
# Make plotting space
par(mar=c(4,4,2,2))
plot(can.rda, scaling=2, display=c("cn", "lc", "sp"), type="n", 
     xlab="RDA1 (16.68 %)", ylab="RDA2 (7.96 %)", main = "Canopy")

# Plot points for sites
spsites.sc <- scores(can.rda, choices=1:2, scaling=2, display="lc")
points(spsites.sc, pch=1, cex=0.5)

# Plot points for species
sp.sc <- scores(can.rda, choices=1:2, scaling=2, display="sp")
points(sp.sc, pch=4, cex=0.5, col="gray50")

# Plot point for geomorphological unit centroids and their labels
unit.names <- c("CH", "LO", "OC", "SU", "TE", "UO")
canunit.sc <- scores(can.rda, choices=1:2, scaling=2, display="cn")
points(canunit.sc, pch=23, bg = "darkgray")
text(canunit.sc[c(2,4,5),], unit.names[c(2,4,5)], cex=0.7, font=2, pos=2, col = "darkgray")
text(canunit.sc[c(3,6),], unit.names[c(3,6)], cex=0.7, font=2, pos=3, col = "darkgray")
text(matrix(canunit.sc[1,], nrow=1, ncol=2), unit.names[c(1)], cex=0.7, font=2, pos=1, col = "darkgray")

# Plot arrows of quantitative explanatory variables and their labels
canenv.sc <- scores(can.rda, choices=1:2, scaling=2, display="bp")
arrows(0,0, canenv.sc[1:5,1], canenv.sc[1:5,2], lty=1, lwd=2.5, length=0.1, col="black")
env.names <- c("Elevation", "Slope", "Rock", "CDI", "Aspect")
text(canenv.sc[c(1,3),], env.names[c(1,3)], cex=0.9, font=2, pos=2)
text(canenv.sc[c(2,4),], env.names[c(2,4)], cex=0.9, font=2, pos=1)
text(matrix(canenv.sc[5,], nrow = 1, ncol = 2), env.names[5], cex=0.9, font=2, pos=4)


# Fig 4b - Understory RDA triplot - Scaling 2 (Correlation triplot)
# Make plotting space
par(mar=c(4,4,2,2))
plot(und.rda, scaling=2, display=c("cn", "lc", "sp"), type="n", 
     xlab="RDA1 (8.22 %)", ylab="RDA2 (7.17 %)", main = "Understory")

# Plot points for sites
spsites.sc <- scores(und.rda, choices=1:2, scaling=2, display="lc")
points(spsites.sc, pch=1, cex=0.5)

# Plot points for species
sp.sc <- scores(und.rda, choices=1:2, scaling=2, display="sp")
points(sp.sc, pch=4, cex=0.5, col="gray50")

# Plot point for geomorphological unit centroids and their labels
unit.names <- c("CH", "LO", "OC", "SU", "TE", "UO")
spunit.sc <- scores(und.rda, choices=1:2, scaling=2, display="cn")
points(spunit.sc, pch=23, bg = "darkgray")
text(spunit.sc[c(2,3),], unit.names[c(2,3)], cex=0.7, font=2, pos=2)
text(spunit.sc[c(1,4,5,6),], unit.names[c(1,4,5,6)], cex=0.7, font=2, pos=4)

# Plot arrows of quantitative explanatory variables and their labels
spenv.sc <- scores(und.rda, choices=1:2, scaling=2, display="bp")
arrows(0,0, spenv.sc[1:5,1], spenv.sc[1:5,2], lty=1, lwd=2.5, length=0.1, col="black")
env.names <- c("Elevation", "Slope", "Rock", "CDI", "Aspect")
text(spenv.sc[1:3,], env.names[1:3], cex=0.9, font=2, pos=4)
text(spenv.sc[4:5,], env.names[4:5], cex=0.9, font=2, pos=2)
```
