# Data Analysis for Block and Meave (2015, Plant Eco & Div)

This report shows the data analysis done for the paper Block and Meave (201X), published in the journal Plant Ecology & Diversity.  
The analysis was reproduced for the last time on March 9th, 2015. I used R version 3.1.2 (2014-10-31) -- "Pumpkin Helmet" and worked in a platform x86_64-apple-darwin13.4.0 (64-bit).

## Redundancy analysis

First, load the packages required for the analysis.

```{r}
library(ade4)
library(vegan)
library(MASS)
library(ellipse)
library(FactoMineR)
```

To load the data, all files must be in the working directory. 


```{r}
structure <- read.csv("str.csv")
env <- read.csv("env.csv")
asp <- read.csv("aspect.csv")
```

I did some modifications to the data. First, I combined all the quantitative variables of the `env` data (i.e., all except the geomorphological unit) with the `asp` data into a single object called `env.quant`. Then, I normalized the data in `env.quant` to allow comparison between variables with very different units. Finally, I added the geomorphological unit column to the normalized quantitative environmental variables to create the data frame `env.ra`, which was used for the redundancy analysis. 

```{r}
env.quant <- cbind(env[,-5],asp)
env.z <- scale(env.quant)
unit <- env[,5]
env.ra <- data.frame(env.z,unit)
```

### Structural data

I did a redundancy analysis on vegetation structure using the environmental data `env.ra` as explanatory variables. I assigned the results to the object `str.rda`.

```{r}
str.rda <- rda(structure~., env.ra, scale=T)
summary(str.rda)
```

I then got the R squared and the adjusted R squared for this analysis.

```{r}
(R2 <- RsquareAdj(str.rda)$r.squared)
(R2adj <- RsquareAdj(str.rda)$adj.r.squared)
```

To test the null hypothesis that no linear relationship exists between the response data and the explanatory variables, I did a permutation analysis with 1000 permutations. 

```{r}
anova.cca(str.rda, step=1000)
```

To make a triplot with scaling 2. 

```{r}
par(mar=c(4,4,2,2))
plot(str.rda, xlab="RDA1 (19.92 %)", ylab="RDA2 (3.51 %)", 
     display=c("cn", "lc", "sp"), type="n", xlim=c(-1.3,1.3))
sites.sc <- scores(str.rda, choices=1:2, scaling=2, display="lc")
points(sites.sc, pch=1, cex=0.5)
va.sc <- scores(str.rda, choices=1:2, scaling=2, display="sp")
text(va.sc, row.names(va.sc), cex=0.9)
env.sc <- scores(str.rda, choices=1:2, scaling=2, display="bp")
arrows(0,0, env.sc[1:5,1], env.sc[1:5,2], lty=1, lwd=2.5, length=0.1)
env.names <- c("Elevation", "Slope", "Rock", "CDI", "Aspect")
text(env.sc[c(1,3,4),], env.names[c(1,3,4)], cex=0.9, font=2, pos=4)
text(env.sc[c(2,5),], env.names[c(2,5)], cex=0.9, font=2, pos=2)
unit.names <- c("CH", "LO", "OC", "SU", "TE", "UO")
unit.sc <- scores(str.rda, choices=1:2, scaling=2, display="cn")
points(unit.sc, pch=23)
text(unit.sc[c(1,2,4,5),], unit.names[c(1,2,4,5)], cex=0.7, font=2, pos=4)
text(unit.sc[c(3,6),], unit.names[c(3,6)], cex=0.7, font=2, pos=3)
```

### Composition data

First I loaded the species data. Again, the file `PA_total.csv` must be in the working directory. To reduce the importance of large abundances, I applied the Hellinger transformation to the species data, as recommended by Borcard et al. (2011, Numerical Ecology in R)
. 
```{r}
spe <- read.csv("PA_total.csv", sep=",", header=T, row.names=1)
spe.he <- decostand(spe,"hellinger")
```

Then I did a redundancy analysis using all the environmental variables as explanatory variables. I tested the null hypothesis with a permutation analysis.

```{r}
spe.rda <- rda(spe.he~., env.ra)
summary.sperda <- summary(spe.rda)
head(summary.sperda)

# Get the R^2 and an adjusted R^2
(spR2 <- RsquareAdj(spe.rda)$r.squared)
(spR2adj <- RsquareAdj(spe.rda)$adj.r.squared)

# Permutation analysis
anova.cca(spe.rda, step=1000) 
```

To make a triplot of the RDA on species data with scaling 2.

```{r}
# Make plotting space
par(mar=c(4,4,2,2))
plot(spe.rda, scaling=2, display=c("cn", "lc", "sp"), type="n", 
     xlab="RDA1 (7.91 %)", ylab="RDA2 (6.01 %)")

# Plot points for sites
spsites.sc <- scores(spe.rda, choices=1:2, scaling=2, display="lc")
points(spsites.sc, pch=1, cex=0.5)

# Plot points for species
sp.sc <- scores(spe.rda, choices=1:2, scaling=2, display="sp")
points(sp.sc, pch=4, cex=0.5, col="gray50")

# Plot arrows of quantitative explanatory variables and their labels
spenv.sc <- scores(spe.rda, choices=1:2, scaling=2, display="bp")
arrows(0,0, spenv.sc[1:5,1], spenv.sc[1:5,2], lty=1, lwd=2.5, length=0.1, col="black")
env.names <- c("Elevation", "Slope", "Rock", "CDI", "Aspect")
text(spenv.sc[1:3,], env.names[1:3], cex=0.9, font=2, pos=4)
text(spenv.sc[4:5,], env.names[4:5], cex=0.9, font=2, pos=1)

# Plot point for geomorphological unit centroids and their labels
unit.names <- c("CH", "LO", "OC", "SU", "TE", "UO")
spunit.sc <- scores(spe.rda, choices=1:2, scaling=2, display="cn")
points(spunit.sc, pch=23)
text(spunit.sc[c(2,3,4,6),], unit.names[c(2,3,4,6)], cex=0.7, font=2, pos=2)
text(spunit.sc[c(1,5),], unit.names[c(1,5)], cex=0.7, font=2, pos=3)
```

### Plot for the paper

This is the code to make the figure for the paper.

```{r, fig.width=6.7, fig.height=10}
par(mfrow=c(2,1))
# Fig 3a - Structure RDA triplot - Scaling 2 (Correlation triplot)
par(mar=c(4,4,2,1))
plot(1, xlab="RDA1 (19.92 %)", ylab="RDA2 (3.51 %)", type="n", 
     xlim=c(-1.3,1.3), ylim=c(-1.1,1.1))
abline(v=0, lty=3); abline(h=0, lty=3); mtext("a", side=3, adj=0, cex=1.25)
sites.sc <- scores(str.rda, choices=1:2, scaling=2, display="lc")
points(sites.sc, pch=1, cex=0.5)
va.sc <- scores(str.rda, choices=1:2, scaling=2, display="sp")
text(va.sc, row.names(va.sc), cex=0.9)
env.sc <- scores(str.rda, choices=1:2, scaling=2, display="bp")
arrows(0,0, env.sc[1:5,1], env.sc[1:5,2], lty=1, lwd=2.5, length=0.1)
env.names <- c("Elevation", "Slope", "Rock", "CDI", "Aspect")
text(env.sc[c(3,4),], env.names[c(3,4)], cex=0.9, font=2, pos=4)
text(env.sc[c(2,5),], env.names[c(2,5)], cex=0.9, font=2, pos=2)
text(env.sc[1,1], env.sc[1,2], env.names[1], cex=0.9, font=2, pos=3)
unit.names <- c("CH", "LO", "OC", "SU", "TE", "UO")
unit.sc <- scores(str.rda, choices=1:2, scaling=2, display="cn")
points(unit.sc, pch=18)
text(unit.sc[c(2,4,5),], unit.names[c(2,4,5)], cex=0.7, font=2, pos=4)
text(unit.sc[c(1,3,6),], unit.names[c(1,3,6)], cex=0.7, font=2, pos=3)

# Fig 3b - Species composition RDA triplot - Scaling 2 (Correlation triplot)
# Make plotting space
par(mar=c(4,4,2,1))
plot(1, type="n", xlim=c(-0.5,1), ylim=c(-0.6,0.8),
     xlab="RDA1 (7.91 %)", ylab="RDA2 (6.01 %)")
abline(v=0, lty=3); abline(h=0, lty=3)
mtext("b", side=3, adj=0, cex=1.25)

# Plot points for sites
spsites.sc <- scores(spe.rda, choices=1:2, scaling=2, display="lc")
points(spsites.sc, pch=1, cex=0.5)

# Plot points for species
sp.sc <- scores(spe.rda, choices=1:2, scaling=2, display="sp")
points(sp.sc, pch=4, cex=0.5, col="gray50")

# Plot arrows of quantitative explanatory variables and their labels
spenv.sc <- scores(spe.rda, choices=1:2, scaling=2, display="bp")
arrows(0,0, spenv.sc[1:5,1], spenv.sc[1:5,2], lty=1, lwd=2.5, length=0.1, col="black")
env.names <- c("Elevation", "Slope", "Rock", "CDI", "Aspect")
text(spenv.sc[1:3,], env.names[1:3], cex=0.9, font=2, pos=4)
text(spenv.sc[4:5,], env.names[4:5], cex=0.9, font=2, pos=1)

# Plot point for geomorphological unit centroids and their labels
unit.names <- c("CH", "LO", "OC", "SU", "TE", "OU")
spunit.sc <- scores(spe.rda, choices=1:2, scaling=2, display="cn")
points(spunit.sc, pch=18)
text(spunit.sc[c(2,3,4,6),], unit.names[c(2,3,4,6)], cex=0.7, font=2, pos=2)
text(spunit.sc[c(1,5),], unit.names[c(1,5)], cex=0.7, font=2, pos=3)
```
