---
title: "Data Analysis in Block and Meave (2017, Plant Eco & Div)"
author: Sebastian Block
output:
  html_document:
    keep_md: true
    theme: flatly
---

This report shows the data analysis presented in the paper Block and Meave (2017), published in the journal Plant Ecology & Diversity.  
The analysis was reproduced for the last time on May 12th, 2017. I used R version 3.3.0 (2016-05-03) -- "Supposedly Educational" and worked in a platform x86_64-apple-darwin13.4.0 (64-bit).

## Community composition and geomorphological units

We used Non-Metric Multidimensional Scaling to explore the floristic similiarity among our plots and to assess if plots in the same geomorphological unit are more similar than what it would be expected based on the geographic distance between them. 

```{r}
### Load required packages
library(ade4)
library(vegan)
library(MASS)
library(ellipse)
library(FactoMineR)

# Load vector containing geomorphological unit of each plot
env <- read.csv("env.csv")
unit <- as.character(env[,5])

### Read composition data and run NMDS for Jaccard and Chao indexes
x <- read.csv("PA_total.csv", sep=",", header=T, row.names=1)

jaccard_dist <- vegdist(x, method = "jaccard")
chao_dist <- vegdist(x, method = "chao")
mds.chao <- metaMDS(x, distance="chao", binary=T ) # Calculate Chao-Jaccard distances

## Calculate spatial distances
Location <- read.csv("Location.csv") # Read plot location data
DistMat <- dist(Location[,4:5])
geog_dist <- as.numeric(DistMat)

# Plots
par(mfrow = c(1,2), mar = c(2.5,2.5,1.2,0.5), cex.axis = 0.6, cex.lab = 0.8, mgp = c(1.5, 0.5, 0))

# NMDS
plot(-1.6:2, -1.6:1.5, xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5),
     type = "n", xlab = "NMDS1", ylab = "NMDS2")
text(mds.chao$points, unit, cex = 0.4, col = "gray20")
mtext("(a)", side = 3, adj = 0, line = 0.2, outer = F)

# Distance vs Chao
plot(geog_dist/1000, chao_dist, ylab = "Chao-Jaccard Dissimilarity Index", 
     xlab = "Distance (km)", cex = 0.4, pch = 19, col = "darkgray")
points(geog_dist/1000, chao_dist, cex = 0.4, col = "gray50")
mtext("(b)", side = 3, adj = 0, line = 0.2, outer = F)
```


## Redundancy analysis

First, load the packages required for the analysis.

```{r}
library(ade4)
library(vegan)
library(MASS)
library(ellipse)
library(FactoMineR)
```

To load the data, all files must be in the working directory. 


```{r}
structure <- read.csv("str.csv") # Vegetation structure data
env <- read.csv("env.csv") # Environmental variables data
asp <- read.csv("aspect.csv") # Aspect data
```

I did some modifications to the data. First, I combined all the quantitative variables of the `env` data (i.e., all variables except the geomorphological unit) with the `asp` data into a single object called `env.quant`. Then, I normalized the data in `env.quant` to allow comparison between variables with very different units. Finally, I added the geomorphological unit column to the normalized quantitative environmental variables to create the data frame `env.ra`, which was used for the redundancy analysis. 

```{r}
env.quant <- cbind(env[,-5],asp)
env.z <- scale(env.quant)
unit <- env[,5]
env.ra <- data.frame(env.z,unit)
```

### Structural data

I did a redundancy analysis on vegetation structure using the environmental data `env.ra` as explanatory variables. I assigned the results to the object `str.rda`.

```{r}
str.rda <- rda(structure~., env.ra, scale=T)
summary(str.rda)
```

I then got the R squared and the adjusted R squared for this analysis.

```{r}
(R2 <- RsquareAdj(str.rda)$r.squared)
(R2adj <- RsquareAdj(str.rda)$adj.r.squared)
```

To test the null hypothesis that no linear relationship exists between the response data and the explanatory variables, I did a permutation analysis with 1000 permutations. 

```{r}
anova.cca(str.rda, step = 1000)
```

To make a triplot with scaling 2. 

```{r}
par(mar=c(4,4,2,2))
plot(str.rda, xlab="RDA1 (19.92 %)", ylab="RDA2 (3.51 %)", 
     display=c("cn", "lc", "sp"), type="n", xlim=c(-1.3,1.3))
sites.sc <- scores(str.rda, choices=1:2, scaling=2, display="lc")
points(sites.sc, pch=1, cex=0.5)
va.sc <- scores(str.rda, choices=1:2, scaling=2, display="sp")
text(va.sc, row.names(va.sc), cex=0.9)
env.sc <- scores(str.rda, choices=1:2, scaling=2, display="bp")
arrows(0,0, env.sc[1:5,1], env.sc[1:5,2], lty=1, lwd=2.5, length=0.1)
env.names <- c("Elevation", "Slope", "Rock", "CDI", "Aspect")
text(env.sc[c(1,3,4),], env.names[c(1,3,4)], cex=0.9, font=2, pos=4)
text(env.sc[c(2,5),], env.names[c(2,5)], cex=0.9, font=2, pos=2)
unit.names <- c("CH", "LO", "OC", "SU", "TE", "UO")
unit.sc <- scores(str.rda, choices=1:2, scaling=2, display="cn")
points(unit.sc, pch=23)
text(unit.sc[c(1,2,4,5),], unit.names[c(1,2,4,5)], cex=0.7, font=2, pos=4)
text(unit.sc[c(3,6),], unit.names[c(3,6)], cex=0.7, font=2, pos=3)
```

### Composition data

#### Canopy

First I loaded the species data. Again, the file `can.csv` must be in the working directory. To reduce the importance of large abundances, I applied the Hellinger transformation to the species data, as recommended by Borcard et al. (2011, Numerical Ecology in R)
. 
```{r}
can <- read.csv("can.csv", sep = ",", header = T, row.names = 1)
can.he <- decostand(can, "hellinger")
```

Then I did a redundancy analysis using all the environmental variables as explanatory variables. I tested the null hypothesis with a permutation analysis.

```{r}
can.rda <- rda(can.he ~ ., env.ra)
summary.canrda <- summary(can.rda)
head(summary.canrda)

# Get the R^2 and an adjusted R^2
(spR2 <- RsquareAdj(can.rda)$r.squared)
(spR2adj <- RsquareAdj(can.rda)$adj.r.squared)

# Permutation analysis
anova.cca(can.rda, step = 1000) 
```

To make a triplot of the RDA on canopy species data with scaling 2.

```{r}
# Make plotting space
par(mar=c(4,4,2,2))
plot(can.rda, scaling=2, display=c("cn", "lc", "sp"), type="n", 
     xlab="RDA1 (19.61 %)", ylab="RDA2 (6.77 %)", main = "Canopy")

# Plot points for sites
spsites.sc <- scores(can.rda, choices=1:2, scaling=2, display="lc")
points(spsites.sc, pch=1, cex=0.5)

# Plot points for species
sp.sc <- scores(can.rda, choices=1:2, scaling=2, display="sp")
points(sp.sc, pch=4, cex=0.5, col="gray50")

# Plot point for geomorphological unit centroids and their labels
unit.names <- c("CH", "LO", "OC", "SU", "TE", "UO")
canunit.sc <- scores(can.rda, choices=1:2, scaling=2, display="cn")
points(canunit.sc, pch=23, bg = "darkgray")
text(canunit.sc[c(2,4,5),], unit.names[c(2,4,5)], cex=0.7, font=2, pos=2, col = "darkgray")
text(canunit.sc[c(3,6),], unit.names[c(3,6)], cex=0.7, font=2, pos=3, col = "darkgray")
text(matrix(canunit.sc[1,], nrow=1, ncol=2), unit.names[c(1)], cex=0.7, font=2, pos=1, col = "darkgray")

# Plot arrows of quantitative explanatory variables and their labels
canenv.sc <- scores(can.rda, choices=1:2, scaling=2, display="bp")
arrows(0,0, canenv.sc[1:5,1], canenv.sc[1:5,2], lty=1, lwd=2.5, length=0.1, col="black")
env.names <- c("Elevation", "Slope", "Rock", "CDI", "Aspect")
text(canenv.sc[c(1,3),], env.names[c(1,3)], cex=0.9, font=2, pos=2)
text(canenv.sc[c(2,4),], env.names[c(2,4)], cex=0.9, font=2, pos=1)
text(matrix(canenv.sc[5,], nrow = 1, ncol = 2), env.names[5], cex=0.9, font=2, pos=4)
```


#### Understorey

First I loaded the species data. Again, the file `und.csv` must be in the working directory. To reduce the importance of large abundances, I applied the Hellinger transformation to the species data, as recommended by Borcard et al. (2011, Numerical Ecology in R)
. 
```{r}
und <- read.csv("und.csv", sep=",", header=T)
und.he <- decostand(und, "hellinger")
```

Then I did a redundancy analysis using all the environmental variables as explanatory variables. I tested the null hypothesis with a permutation analysis.

```{r}
und.rda <- rda(und.he ~ ., env.ra)
summary.undrda <- summary(und.rda)
head(summary.undrda)

# Get the R^2 and an adjusted R^2
(spR2 <- RsquareAdj(und.rda)$r.squared)
(spR2adj <- RsquareAdj(und.rda)$adj.r.squared)

# Permutation analysis
anova.cca(und.rda, step=1000) 
```

To make a triplot of the RDA on species data with scaling 2.

```{r}
# Make plotting space
par(mar=c(4,4,2,2))
plot(und.rda, scaling=2, display=c("cn", "lc", "sp"), type="n", 
     xlab="RDA1 (9.76 %)", ylab="RDA2 (6.67 %)", main = "Understorey")

# Plot points for sites
spsites.sc <- scores(und.rda, choices=1:2, scaling=2, display="lc")
points(spsites.sc, pch=1, cex=0.5)

# Plot points for species
sp.sc <- scores(und.rda, choices=1:2, scaling=2, display="sp")
points(sp.sc, pch=4, cex=0.5, col="gray50")

# Plot point for geomorphological unit centroids and their labels
unit.names <- c("CH", "LO", "OC", "SU", "TE", "UO")
spunit.sc <- scores(und.rda, choices=1:2, scaling=2, display="cn")
points(spunit.sc, pch=23, bg = "darkgray")
text(spunit.sc[c(2,3),], unit.names[c(2,3)], cex=0.7, font=2, pos=2)
text(spunit.sc[c(1,4,5,6),], unit.names[c(1,4,5,6)], cex=0.7, font=2, pos=4)

# Plot arrows of quantitative explanatory variables and their labels
spenv.sc <- scores(und.rda, choices=1:2, scaling=2, display="bp")
arrows(0,0, spenv.sc[1:5,1], spenv.sc[1:5,2], lty=1, lwd=2.5, length=0.1, col="black")
env.names <- c("Elevation", "Slope", "Rock", "CDI", "Aspect")
text(spenv.sc[1:3,], env.names[1:3], cex=0.9, font=2, pos=4)
text(spenv.sc[4:5,], env.names[4:5], cex=0.9, font=2, pos=2)
```


### Plot for the paper

This is the code to make the figure for the paper.

```{r, fig.width=6.7, fig.height=10}
par(mfrow=c(2,1))

# Figure 4a - Canopy
# Make plotting space
par(mar=c(4,4,2,2))
plot(can.rda, scaling=2, display=c("cn", "lc", "sp"), type="n", 
     xlab="RDA1 (19.61 %)", ylab="RDA2 (6.77 %)", main = "Canopy")

# Plot points for sites
spsites.sc <- scores(can.rda, choices=1:2, scaling=2, display="lc")
points(spsites.sc, pch=1, cex=0.5)

# Plot points for species
sp.sc <- scores(can.rda, choices=1:2, scaling=2, display="sp")
points(sp.sc, pch=4, cex=0.5, col="gray50")

# Plot point for geomorphological unit centroids and their labels
unit.names <- c("CH", "LO", "OC", "SU", "TE", "UO")
canunit.sc <- scores(can.rda, choices=1:2, scaling=2, display="cn")
points(canunit.sc, pch=23, bg = "darkgray")
text(canunit.sc[c(2,4,5),], unit.names[c(2,4,5)], cex=0.7, font=2, pos=2, col = "darkgray")
text(canunit.sc[c(3,6),], unit.names[c(3,6)], cex=0.7, font=2, pos=3, col = "darkgray")
text(matrix(canunit.sc[1,], nrow=1, ncol=2), unit.names[c(1)], cex=0.7, font=2, pos=1, col = "darkgray")

# Plot arrows of quantitative explanatory variables and their labels
canenv.sc <- scores(can.rda, choices=1:2, scaling=2, display="bp")
arrows(0,0, canenv.sc[1:5,1], canenv.sc[1:5,2], lty=1, lwd=2.5, length=0.1, col="black")
env.names <- c("Elevation", "Slope", "Rock", "CDI", "Aspect")
text(canenv.sc[c(1,3),], env.names[c(1,3)], cex=0.9, font=2, pos=2)
text(canenv.sc[c(2,4),], env.names[c(2,4)], cex=0.9, font=2, pos=1)
text(matrix(canenv.sc[5,], nrow = 1, ncol = 2), env.names[5], cex=0.9, font=2, pos=4)


# Fig 4b - Understory RDA triplot - Scaling 2 (Correlation triplot)
# Make plotting space
par(mar=c(4,4,2,2))
plot(und.rda, scaling=2, display=c("cn", "lc", "sp"), type="n", 
     xlab="RDA1 (9.76 %)", ylab="RDA2 (6.67 %)", main = "Understorey")

# Plot points for sites
spsites.sc <- scores(und.rda, choices=1:2, scaling=2, display="lc")
points(spsites.sc, pch=1, cex=0.5)

# Plot points for species
sp.sc <- scores(und.rda, choices=1:2, scaling=2, display="sp")
points(sp.sc, pch=4, cex=0.5, col="gray50")

# Plot point for geomorphological unit centroids and their labels
unit.names <- c("CH", "LO", "OC", "SU", "TE", "UO")
spunit.sc <- scores(und.rda, choices=1:2, scaling=2, display="cn")
points(spunit.sc, pch=23, bg = "darkgray")
text(spunit.sc[c(2,3),], unit.names[c(2,3)], cex=0.7, font=2, pos=2)
text(spunit.sc[c(1,4,5,6),], unit.names[c(1,4,5,6)], cex=0.7, font=2, pos=4)

# Plot arrows of quantitative explanatory variables and their labels
spenv.sc <- scores(und.rda, choices=1:2, scaling=2, display="bp")
arrows(0,0, spenv.sc[1:5,1], spenv.sc[1:5,2], lty=1, lwd=2.5, length=0.1, col="black")
env.names <- c("Elevation", "Slope", "Rock", "CDI", "Aspect")
text(spenv.sc[1:3,], env.names[1:3], cex=0.9, font=2, pos=4)
text(spenv.sc[4:5,], env.names[4:5], cex=0.9, font=2, pos=2)
```
